---
title: "Assignment1"
author: "Mark Fernandes"
date: "08/04/2020"
output: html_document
---

# __CRUK Online Introduction to R course__

1) You need to generate an X-axis variable that incrememtally increases and represents the sampling at 5 day intervals for about a year (365 days).    
Your friendly neighbourhood statistician has suggested that there should be an R function to do that. What is the function and how do you find out about it and what is the code you will use to create the X-axis vector? Check the vector.
```{r}

```

2) You have been asked to plot some base R graphs of counts data. The investigator has requested that certain symbols be used for each dataset being plotted.    
What command would you use to find out the parameter which sets this for the plot.default() command and what is the parameters name?
```{r}

```

3) Your colleague has supplied you with the following table of data (number of cells per sample volume):   

| Day | LineA | LineB | LineC |
| --  | ---   | ---   | ---   |
| 1   |   4   |    5  |   14   |
| 2   |   9   |   17  |   16   |
| 3   |   7   |   22  |   10  |
| 4   |   12   |  20   |   14  |
| 5   |   23    |   24  |   20 |
| 6   |   8    |  18  |   12    |

Create some R vectors to hold this data and provide summary statisics for number of cells for each cell line. Plot some base R graphs if you like. Describe the data. 
```{r}

```

You are then provided with assay data that states that LineA had an activity of 4.2 per cell, LineB of 3.4 and LineC of 1.3.    
Use R to calculate the activities of each sample on each day and provide summary statistics of activity for each line.    
```{r}

```



