
---
title: "Assignment 4"
---

The week's assignment will test your ability to manipulate the METABRIC data set
by changing the values of existing columns or adding new columns by computing
new variables from existing ones.

We are expecting you to use the 5 main dplyr 'verb' functions: `select()`,
`filter()`, `arrange()`, `mutate()` and `summarize()`. Please use the pipe
operator, `%>%`, in cases where more than one operation is required to achieve
the outcome requested.

```{r message = FALSE, warning = FALSE}
library(tidyverse)
metabric <- read_csv("metabric_clinical_and_expression_data.csv")
```

---

**1. Investigate the subset of long-surviving breast cancer patients that didn't receive chemo or radiotherapy**

First obtain the subset of patients that received neither chemotherapy or
radiotherapy and survived for more than 20 years.

```{r}

```

Now look at the breakdown of these patients in terms of ER status. Count the
numbers of ER positive and ER negative patients in this subset. Calculate the
proportion that are ER positive.

```{r}

```

What does this tell us? Calculate the proportion of ER positive patients in the
whole cohort by way of a comparison.

```{r}

```

**Extension**

Create a contingency table of the numbers of ER positive and ER negative
patients in the two groups (untreated, long-surviving patients and all others)
and perform a Chi-squared test or a Fisher's exact test (whichever is most
appropriate) to determine if there is a significant difference.

```{r}

```

_Hint: look at the help page for the function for the Chi-squared or Fisher's
exact test to see what form the contingency table needs to take_

---

**2. Which patients have spent the largest proportion of their lives dealing with breast cancer?**

```{r}

```

---

**3. Convert the expression values for each of the genes into standardized z-scores**

Some genes are generally expressed at higher levels than others. This can make
comparisons of changes between groups for a set of genes somewhat difficult,
particularly if the expression for those genes are on very different scales. The
expression values in our METABRIC are on a log~2~ scale which helps to reduce
the range of values but another method for representing expression measurements
is to standardize these to produce z-scores.

Standardization of a set of measurements involves subtracting the mean from each
and dividing by the standard deviation. This will produce values with a mean of
0 and a standard deviation of 1.

Create a modified version of the `metabric` data frame containing a new column
with the standardized expression values (z-scores) for the ESR1 gene.

```{r}

```

Check that you've done this correctly by calculating the mean and standard
deviation of your new z-score variable.

```{r}

```

Add another column to your modified `metabric` data frame containing a z-score
for GATA3 and then create a scatter plot of the z-scores of GATA3 against ESR1.
Modify your plot to facet by the PAM50 classification.

```{r}

```

**Extension 1**

Standardize the expression values for all genes in a single operation using an
anonymous function, overwriting their original values, and round the resulting
values to 3 significant figures.

```{r}

```

Check you have done this correctly by computing the mean and standard deviation for
each column, again using a single operation.

```{r}

```

To be doubly certain, you could also download the expression z-score values for
these genes from cBioPortal.

**Extension 2**

Create a plot comparing the distribution of standardized expression values for
TP53 against a normal distribution.

```{r}

```

_Hint: use `stat_function()`_

---

**4. Which Star Wars characters need to go on a diet?**

Compute the body mass index (BMI) of characters in the `starwars` tibble.

```{r}

```

Filter for human characters that are overweight (BMI > 25) and display in
decreasing order of BMI.

```{r}

```

