---
title: "Assignment_runthrough"
author: "Mark Fernandes"
date: "16/04/2020"
output:
  word_document: default
  html_document: default
---




Now on with this week's exercises.

# Using R as a calculator

1. Convert the following temperatures given in degrees Fahrenheit to Celsius:
45, 96, 451

```{r}
#
#
#
#
#
(45 - 32) * 5/9 
(96 - 32) * 5/9 
(451 - 32) * 5/9 
```

Hint: just do what you'd normally do if you can't remember the formula for
converting between Celsius and Fahrenheit (Google in my case).

If you like, you can experiment with getting your R code right in the Console
window first and then copy it into the code chunk above when you're happy with
it. It's not crucial and getting it wrong in the R markdown is no big deal.
You can always fix any problems (the most likely being forgetting to use
parentheses or brackets in the right place) and re-run your code using the
green arrow/triangle icon.

Check you've got the right answer by finding a web page with a handy
conversion tool.

2. Similarly, convert the following temperatures in degrees Celsius to
Fahrenheit: -65, 100, 20

```{r}
#
#
#
#
-65 * 9/5 + 32
100 * 9/5 + 32
20 * 9/5 + 32
```










# Generating sequence vectors

3. Generate a sequence of numbers representing the days at which you take a
measurement or a sample at 5-day intervals for about a year.





```{r}
1:73 * 5
```



Your friendly neighbourhood statistician has suggested that there should be an R
function to do that. What is the function and how do you find out about it and
what is the code you will use to create the sequence? Check the resulting vector.



```{r}
#
#
#
#
#
#
#
#
??sequence
sample_days <- seq(5, 365, by = 5)
sample_days
```

# Looking at types of objects

4. Run the code below and use the typeof() and/or class() function (check it's
help page) and see how R treats each newly-created vector?

```{r}
num_char <- c(1, 2, 3, "a")
num_logical <- c(1, 2, 3, TRUE)
char_logical <- c("a", "b", "c", TRUE)
tricky <- c(1, 2, 3, "4")
#
#
#
#
#
# Answers to the type coercion question below
#
#
#
#
#
typeof(num_char)
class(num_char)
num_char
typeof(num_logical)
class(num_logical)
num_logical
typeof(char_logical)
class(char_logical)
char_logical
typeof(tricky)
class(tricky)
tricky
```

Create a new code chunk to test each of the vectors in a separate block. You
can do this by using the 'Insert' menu just at the top of the pane for this
markdown file and selecting R for an R code chunk, or by using the keyboard
shortcut (on a Mac this is cmd-alt-i).

You should find that R coerces the data to a lowest common denominator - can you
work out the hierarchy?
```{r}
# logical → numeric → character ← logical 
```


# Plotting data

5. You have been asked to plot a graph of counts data measured over several
days. The Principal Investigator has requested that certain symbols be used for
each dataset being plotted (he's a bit like that!). What command would you use
to find out the parameter which sets this for the plot.default command and what
is the parameter's name?

```{r}
days <- c(1, 2, 4, 6, 8, 12, 16)
counts <- days ^ 2 + rnorm(days, mean = days)

#
#
#
#
#
#
#
#
# add your code here
?rnorm
rnorm(days, mean = days)
#
help("plot.default")
?plot.default
#
plot.default(days,counts, pch=6)
```

Check out what we did in the above example for getting some example counts
data points. Can you make sense of what is going on here? Look at the help
page for the `rnorm` function.

Our counts data don't really look like counts as they are not whole numbers.
Find the function in R that can round these up or down to the closest whole
number and apply it in the above code chunk.
```{r}
??round
round_days <- round(counts)
round_days
plot.default(days,round(counts), pch=6)
```


# Exploring and summarizing data

6. Your colleague has supplied you with the following table of data (number of
cells per sample volume):

| Day | LineA | LineB | LineC |
| --- | ----- | ----- | ----- |
|  1  |    4  |    5  |   14  |
|  2  |    9  |   17  |   16  |
|  3  |    7  |   22  |   10  |
|  4  |   12  |   20  |   14  |
|  5  |   23  |   24  |   20  |
|  6  |    8  |   18  |   12  |

Create some R vectors to hold this data and provide summary statisics for number
of cells for each cell line. Plot some base R graphs if you like. Describe the
data.

```{r}
day <- c(1,2,3,4,5,6)
line_a <- c(4,9,7,12,23,8)
line_b <- c(5,17,22,20,24,18)
line_c <- c(14,16,10,14,20,12)
summary(line_a)
summary(line_b)
summary(line_c)
#
```

You are then provided with assay data that states that LineA had an activity of
4.2 per cell, LineB an activity of 3.4 and LineC of 1.3.

Use R to calculate the activities of each sample on each day and provide summary
statistics of activity for each line.

```{r}
act_line_a <- line_a * 4.2
act_line_b <- line_b * 3.4
act_line_c <- line_c * 1.3
act_line_a
act_line_b
act_line_c
summary(act_line_a)
summary(act_line_b)
summary(act_line_c)
plot.default(day,act_line_a,pch=6)
plot.default(day,act_line_b,pch=7)
plot.default(day,act_line_c,pch=8)
#
# Plot with multiple lines (Yes I had to look it up in help as I don't tend to use base R plot very often)

# plot the first curve by calling plot() function
# First curve is plotted
 plot(day, act_line_a, type="o", col="blue", pch="o", lty=1 )

# Add second curve to the same plot by calling points() and lines()
# Use symbol '*' for points.
 points(day, act_line_b, col="red", pch="*")
 lines(day, act_line_b, col="red",lty=2)

# Add Third curve to the same plot by calling points() and lines()
# Use symbol '+' for points.
 points(day, act_line_c, col="dark green",pch="+")
 lines(day, act_line_c, col="dark green", lty=3)
```

# Creating a report for your assignment

Click on the 'Knit' menu at the top of this file and select either whichever
option you prefer to create an HTML, PDF or Word document version of your
assignment. This will run all the code chunks and "knit" the resulting
results with the surrounding text to produce a report.

