---
title: "Week 4 -- Live Coding script"
---

```{r setup}
options(tibble.print_max = 1, tibble.print_min = 1, max.print=10, 
        tibble.max_extra_cols=1)
knitr::opts_chunk$set(eval=FALSE)
```


## Intro 

* So far:
    - introduced basic R for working with data
    - introduced `ggplot2` - easier for plotting than base R
* `ggplot2` is part of tidyverse.
* History of tidyverse v base R - it's much easier!!!!
* Today:
    - core tidyverse package **dplyr**
    - **pipe**
    - facetting *maybe not in this session*??

Let's start by loading the the tidyverse package and the metabric data

```{r, message=FALSE}
library(tidyverse)
metabric <- read_csv("data/metabric_clinical_and_expression_data.csv")
```

## Filtering rows

##### [DON'T SAY **"SELECT"**] Sometimes we want to ***keep*** only certain rows from a data frame

In base R we use `[, ]`
If we wanted the first five rows:

```{r}
metabric[1:5,]
```

More commonly we want to **filter** based on the contents of column

##### e.g. last week we wanted only deceased patients for one of the plots:

```{r, eval=FALSE}
metabric$Survival_status == "DECEASED"

metabric[metabric$Survival_status=="DECEASED", ]
```

```{r, echo=FALSE}
metabric[metabric$Survival_status=="DECEASED", ]
```

##### This `[]` syntax is not really intuitive

`dplyr` provides a simple function that allows us to **filter** a table in this 
way

and it's called...**`filter`**

```{r}
filter(metabric, Survival_status=="DECEASED")

deceased <- filter(metabric, Survival_status=="DECEASED")
```

##### Talk about grammar of tidyverse - **verbs**
* Have a common structure and behaviour:
    * 1st argument is ALWAYS a data frame to work on (`ggplot` from last week)
    * Additional arguments describe the operation
    * Variables names - column headers - no quotes
    * output is ALWAYS a new data frame
    
##### Other conditionals...if instead we'd used the `Vital_status` column

```{r}
unique(metabric$Vital_status)
```

#### `==` is won't work as there are two catergories for deceased, we could use `%in%`

```{r}
filter(metabric, Vital_status%in%c("Died of Disease", "Died of Other Causes"))
```

##### It would be easier to just look for "Died" - `stringr` (it was loaded in tidyverse)
##### look in the packages

```{r, eval=FALSE}
str_starts(metabric$Vital_status, "Died")

filter(metabric, str_starts(Vital_status, "Died"))
```

```{r, echo=FALSE}
filter(metabric, str_starts(Vital_status, "Died"))
```

Here you can talk about tidyverse v base - c.f gsub/grep v strsplit argument postion

##### filter by multiple conditions

e.g were still alive and survived > 10 years

```{r}
filter(metabric, Survival_status == "LIVING", Survival_time > 120)
```

...same as...

```{r}
filter(metabric, Survival_status == "LIVING" & Survival_time > 120)
```

##### **OR**

ER or PR positive...

```{r, eval=FALSE}
filter(metabric, ER_status == "Positive" | PR_status == "Positive")

filter(metabric, Survival_status == "LIVING" & Survival_time > 120, ER_status == "Positive" | PR_status == "Positive")
```

```{r, echo=FALSE}
filter(metabric, Survival_status == "LIVING" & Survival_time > 120, ER_status == "Positive" | PR_status == "Positive")
```

## Selecting columns

We can't really see the ER and PR status columns to check things, we'd like to 
change the table to select the columns we're interested in and check that we 
only have the patients of interest

```{r}
patients_of_interest <- filter(metabric, 
                               Survival_status == "LIVING" & Survival_time > 120, 
                               ER_status == "Positive" | PR_status == "Positive")
patients_of_interest[,c("Patient_ID", 
                        "Survival_status", 
                        "Survival_time", 
                        "ER_status", 
                        "PR_status")]
```

Again the `[]` syntax is a bit clunky and so `dplyr` provides a simple function 
for selecting columns and it's called...
**`select`**

```{r eval=FALSE}
select(patients_of_interest, Patient_ID)

select(patients_of_interest, Patient_ID, Survival_status, Survival_time, ER_status, PR_status)
```

```{r echo=FALSE}
select(patients_of_interest, Patient_ID, Survival_status, Survival_time, ER_status, PR_status)
```

The order of the columns is how they'll come out

```{r}
select(patients_of_interest, Patient_ID, ER_status, PR_status, Survival_status, Survival_time)
```

dplyr provides helper functions for choosing columns - see the cheatsheet

```{r, eval=FALSE}
select(metabric, contains("Status"))

select(metabric, ends_with("Status"))
```

```{r, echo=FALSE}
select(metabric, ends_with("Status"))
```

##### Exclude columns with **`-`**

```{r}
select(metabric, -Patient_ID)
```

## Chaining operations - the PIPE - `%>%`

Lets say we want to create a new table that:  

* contains living patients that survive > 10 years   
* Only has a few columns of interest
  
```{r}
patients <- filter(metabric, Survival_status == "LIVING", Survival_time > 120)
patient_details <- select(patients, Patient_ID, Survival_status)
patient_details
```

##### The intermediate table **patients** is not actually needed

##### We could reuse the same object

```{r}
patient_details <- filter(metabric, Survival_status == "LIVING", Survival_time > 120)
patient_details <- select(patient_details, Patient_ID, Survival_status)
patient_details
```

##### or (HORROR) nest the commands

```{r}
patient_details <- select(filter(metabric, Survival_status == "LIVING", Survival_time > 120), 
                          Patient_ID, Survival_status)
patient_details
```

##### this is where the pipe comes in...

##### %>%

Takes the output on the left and puts into the first argument of the command on 
the right

```{r}
filter(metabric, Survival_status == "LIVING")
```

same as

```{r}
metabric %>%
  filter(Survival_status == "LIVING")
```

but then we can extend the chain of functions with more pipes


```{r}
metabric %>%
  filter(Survival_status == "LIVING") %>% 
  select(Patient_ID, Survival_time)
```

The **output** data frame is **flowing** through the pipe to become the 
**input** for the next command.

There is no need to restate `metabric`, in fact this would be wrong...

* Shortcut for the pipe is <Ctrl>+<Shift>+M (<CMD>+<SHIFT>+M on a MAC)

##### The `%>%` must be on the end of the line so that R knows to continue

```{r, eval=FALSE}
patients_of_interest <- metabric %>%
  filter(Survival_status == "LIVING")
  %>% filter(Survival_time > 120)
```

## Creating new columns

The third thing that we may want to do is to creat new columns in our data frame, often based
on the contents of other columns. E.g. ....

In base `R` we could do something like:

```{r}
metabric$Deceased <- metabric$Survival_status == "DECEASED"
```

but it would be great if we had a function to this which would fit with the pipe

`dplyr` provides a simple function that allows us to **create** a new column in 
a table
and it's called...**`mutate`** ?!?!?
Basically, the concept is that we will be taking the contents of existing 
column(s) and *mutating* it to create the contents of new column

Once again the syntax follows the standard `tidyverse` patterns: data first, 
column names can be used directly

```{r}
mutate(metabric, Deceased = Survival_status == "DECEASED")
```

and we can fit it in with the pipe:

```{r}
metabric %>%
  filter(Survival_time > 240) %>% 
  mutate(Deceased = Survival_status == "DECEASED") %>% 
  select(Patient_ID, Survival_status, Deceased, Survival_time)
```

NOTE THE DIFFERENCE BETWEEN TEH `=` and the `==`

You *can* also alter existing columns

```{r}
metabric %>%
  filter(Survival_time > 240) %>% 
  mutate(Deceased = Survival_status == "DECEASED") %>% 
  mutate(Survival_time = round(Survival_time)) %>% 
  select(Patient_ID, Survival_status, Deceased, Survival_time)
```

##### You can mutate multiple columns at once

```{r}
metabric %>% 
  select(Patient_ID, Survival_status, Vital_status)

metabric %>% 
  mutate(Survival_status = as.factor(Survival_status), Vital_status = as.factor(Vital_status)) %>% 
  select(Patient_ID, Survival_status, Vital_status)
```

##### Another option is to use **`mutate_at`**

```{r}
metabric %>% 
  mutate_at(vars(Survival_status, Vital_status), as.factor) %>% 
  select(Patient_ID, ends_with("_status"))
```

##### See help and DOCS for more (including anonymous functions)

## Sorting using **`arrange()`**

mention briefly

```{r}
metabric %>% 
  arrange(Age_at_diagnosis)
```

## Summary values using **`sumarise`**

We've seen at various ways to get summary values from columns in base R

```{r}
mean(metabric$ESR1)
summary(metabric)
```

It'd be handy to have a function we could use within a pipe.

`dplyr` provides a simple function that allows us to **summarise** the contents 
of a column and it's called...**`summarise`**

Let's say we want the mean survival time for deceased patients:

```{r}
metabric %>%
  filter(Survival_status == "DECEASED") %>% 
  summarise(Survival_time_mean = mean(Survival_time))
```

##### We can do more than one summary at once

```{r}
metabric %>%
  filter(Survival_status == "DECEASED") %>% 
  summarise(Survival_time_mean = mean(Survival_time), Survival_time_sd = sd(Survival_time))
```

##### More than one column

```{r}
metabric %>%
  filter(Survival_status == "DECEASED") %>% 
  summarise(Survival_time_mean = mean(Survival_time), ESR1_mean = mean(ESR1))
```

##### As with mutate there are **`summarise_at`** etc

## Facetting

Split a plot into separate plots based on the contents of a column


```{r}
ggplot(data = metabric, mapping = aes(x = GATA3, y = ESR1)) +
  geom_point(mapping=aes(colour = Cancer_type))
```


```{r}
ggplot(data = metabric, mapping = aes(x = GATA3, y = ESR1)) +
  geom_point() +
  facet_wrap(~ Cancer_type)
```

```{r}
ggplot(data = metabric, mapping = aes(x = GATA3, y = ESR1)) +
  geom_point(mapping=aes(colour = Cancer_type)) +
  facet_wrap(~ Cancer_type)
```


