---
title: "Week 5 -- Live Coding script"
---

```{r setup}
options(tibble.print_max = 1, tibble.print_min = 1, max.print=10, 
        tibble.max_extra_cols=1)
knitr::opts_chunk$set(eval=FALSE)
```


## Intro 

* So far:
    - introduced basic R for working with data
    - introduced `ggplot2` - easier for plotting than base R
* `ggplot2` is part of tidyverse.
* Last week:
    - dplyr - mutate etc.
    - The pipe
* Today:
    - grouping to make summarise more powerful
    - joining tables
    - some ggplot customisations
    
Let's start by loading the the tidyverse package and the metabric data

```{r, message=FALSE}
library(tidyverse)
metabric <- read_csv("data/metabric_clinical_and_expression_data.csv")
```

## Grouping

##### Last week we looked at summarise:

```{r}
metabric %>% 
  summarise(Mean_ESR1 = mean(ESR1))
```

##### What if we want to look at the mean expression of ESR1 in patients with different Tumour stages

We could use a filter...

```{r}
metabric %>% 
  filter(Tumour_stage==1) %>% 
  summarise(Mean_ESR1_positive = mean(ESR1))
metabric %>% 
  filter(Tumour_stage==2) %>% 
  summarise(Mean_ESR1_negative = mean(ESR1))
```

###### etc.....

#### But dplyr provides a better way - grouping with `group_by`. It's a bit like facetting...

```{r}
metabric %>% 
  group_by(Tumour_stage) %>% 
  summarise(Mean_ESR1 = mean(ESR1))
```

##### How is it doing this? **show the figure**

##### We could even group by multiple categories - say we want to split by ER and HER2 status...

```{r}
metabric %>% 
  group_by(ER_status, HER2_status) %>% 
  summarise(Mean_ESR1 = mean(ESR1))
```

###### As before we can get multiple summary stats in one table

```{r}
metabric %>% 
  group_by(ER_status, HER2_status) %>% 
  summarise(Mean_ESR1 = mean(ESR1), Mean_GATA3 = mean(GATA3))
```

##### It's often useful to have a count of the number of patients in each group

**`n()`**

```{r}
metabric %>% 
  group_by(ER_status, HER2_status) %>% 
  summarise(Mean_ESR1 = mean(ESR1), Mean_GATA3 = mean(GATA3), N_Patients = n())
```

###### In fact quite often this is all we want to know
###### e.g how many in each PAM50 class??

```{r}
metabric %>% 
  group_by(PAM50) %>% 
  summarise(N_Samples=n())
```

##### Dplyr provides a handy wrapper function for this - **`count`**

```{r}
count(metabric, PAM50)
```

##### `count` also does **`ungroup`**

##### Lets say we want to plot mean ER v mean GAT3 for different groups

##### and we want to fix the labels Positive and Negative in the plot

```{r}
metabric %>% 
  group_by(ER_status, HER2_status) %>% 
  summarise(Mean_ESR1 = mean(ESR1), Mean_GATA3 = mean(GATA3)) %>% 
  mutate(ER_status = str_replace_all(ER_status,c("Negative" = "ER-", "Positive" = "ER+")))
```

```{r}
metabric %>% 
  group_by(ER_status, HER2_status) %>% 
  summarise(Mean_ESR1 = mean(ESR1), Mean_GATA3 = mean(GATA3)) %>% 
  ungroup() %>% 
  mutate(ER_status = str_replace_all(ER_status, c("Negative" = "ER-", "Positive" = "ER+")))
```

## Unique rows

##### Sometimes we want to know how many unique values there are

A new data set:

```{r}
mutations <- read_csv("data/metabric_mutations.csv")
```

Let's just use a couple of columns for this exercise

```{r}
mutations <- mutations %>% 
  select(Patient_ID, Gene, Position = Start_Position)  ### LOOK WE CAN RENAME IN SELECT!!!!
mutations
```

##### Let's say we just want a table that shows all mutated genes in each patient

##### i.e. one row for each patient/gene combo

```{r}
mutations %>% 
  distinct(Patient_ID, Gene)
```

## Joining

##### Often we want to join two (or more) tables together based on a common column e.g. ...

##### A few different commands depending on the rules we want to use to do the join

##### The simplest join - **`left_join`**

##### Lets work with some simpler tables

```{r}
band_members
```

```{r}
band_instruments
```

###### Let's join them together

```{r}
left_join(band_members, band_instruments, by = "name")
```

##### Keeps the rows of the lefthand table and adds mathching from the right - Point out the `NA`

##### We can do this in the pipe
##### You don't actually have to give it the `by`

```{r}
band_members %>% 
  left_join(band_instruments)
```

###### There is an opposite - `right_join` - that keeps the right hand table

#### `inner_join` - returns the intersection, i.e. rows in both

```{r}
inner_join(band_members, band_instruments)
```

#### `full_join` - returns the union, i.e. all rows

```{r}
full_join(band_members, band_instruments)
```

### Sometimes the column headers don't match

###### This table has `artist` instead of `name`

```{r}
band_instruments2
```

```{r}
left_join(band_members, band_instruments2, by = c("name" = "artist"))
```

###### What happens if there are multiple matches??? 

Paul was in two bands...

```{r}
band_members2 <- tibble(
    name = c("Mick",   "John",    "Paul",    "Paul"),
    band = c("Stones", "Beatles", "Beatles", "Wings")
)
```

###### Now there are two rows in the left with the same match on the right...

```{r}
left_join(band_members2, band_instruments)
```

###### Think about e.g. the table of mutations, we might want to add gene annotations....

###### What if there are two matches on the right? e.g. signalling pathways

And Paul played two instruments ...

```{r}
band_instruments3 <- tibble(
    name =  c("John",   "Paul", "Paul",   "Keith"),
    plays = c("guitar", "bass", "guitar", "guitar")
)
band_instruments3
```

```{r}
left_join(band_members2, band_instruments3)
```

##### Sometimes we want to join on multiple columns - e.g. Name and Surname

```{r}
band_members4 <- tibble(
    forename = c("Mick",   "John",    "Paul",      "Mick", "John"),
    surname =  c("Jagger", "Lennon",  "McCartney", "Avory", "Squire"),
    band =     c("Stones", "Beatles", "Beatles",   "Kinks", "Roses")
)

band_instruments4 <- tibble(
    forename = c("John",   "Paul",      "Keith",    "John", "Mick"),
    surname =  c("Lennon", "McCartney", "Richards", "Squire", "Avory"),
    plays =    c("guitar", "bass",      "guitar",   "guitar", "drums")
)
```

Just add both column names to the `by` as a vector

```{r}
full_join(band_members4, band_instruments4, by = c("forename", "surname"))
```
##### In fact `_join`s will use all matching columns if not given `by`

```{r}
full_join(band_members4, band_instruments4)
```

###### More in the docs about what happens if column names clash and filtering joins

###### and a more practical example using the metabric data

## Customising plots

### Titles and labels

```{r}
ggplot(metabric, mapping = aes(x=GATA3, y=ESR1)) +
  geom_point(mapping = aes(colour=ER_status))
```

###### The axis labels are okay, but maybe we want them to be clearer
###### and it would be nice if the legend lost the `_`

The expression values are log2

```{r}
ggplot(metabric, mapping = aes(x=GATA3, y=ESR1)) +
  geom_point(mapping = aes(colour=ER_status)) +
  labs(x = "log2(GATA3 expression)", 
       y = "log2(ESR1 expression)",
       colour = "ER Status")
```

##### and how about a plot title....


```{r}
ggplot(metabric, mapping = aes(x=GATA3, y=ESR1)) +
  geom_point(mapping = aes(colour=ER_status)) +
  labs(x = "log2(GATA3 expression)", 
       y = "log2(ESR1 expression)",
       colour = "ER Status",
       title = "mRNA expression in METABRIC data set")
```

can also add subtitles

### Changing the axes - SCALES

##### all the aesthetics are "scales" - either continuous or discrete

ggplot is adding the scale legends/axes for us, but we can add them manually

```{r}
ggplot(metabric, mapping = aes(x=GATA3, y=ESR1)) +
  geom_point(mapping = aes(colour=ER_status)) +
  scale_x_continuous() +
  scale_y_continuous() +
  scale_colour_discrete()
```

##### we only need to do this if we are going to modify aspects of their display

##### let's change the x-axis a bit - LOOK AT THE HELP 

##### So, modify the breaks to be 5 - 15 at intervals of 2.5

```{r}
ggplot(metabric, mapping = aes(x=GATA3, y=ESR1)) +
  geom_point(mapping = aes(colour=ER_status)) +
  scale_x_continuous(breaks = seq(5, 15, by = 2.5))
```

##### Note the limits - they have not changed

```{r}
ggplot(metabric, mapping = aes(x=GATA3, y=ESR1)) +
  geom_point(mapping = aes(colour=ER_status)) +
  scale_x_continuous(breaks = seq(5, 15, by = 2.5), limits = c(5, 14))
```

###### More mods in the documention

### Colour - a discrete scale (note that can be continuous)

##### scale_colour_discrete automatically sets the colours.
##### To manually alter them - `scale_colour_manual` - LOOK AT HELP

```{r}
ggplot(metabric, mapping = aes(x=GATA3, y=ESR1)) +
  geom_point(mapping = aes(colour=ER_status)) +
  scale_colour_manual(values = c("dodgerblue2", "firebrick2"))
```

##### When we have lots of colours it becomes tiresome to choose them all ourselves


```{r}
ggplot(metabric, mapping = aes(x=GATA3, y=ESR1)) +
  geom_point(mapping = aes(colour=PAM50))
```

##### Introduce colour brewer

```{r}
ggplot(metabric, mapping = aes(x=GATA3, y=ESR1)) +
  geom_point(mapping = aes(colour=PAM50)) +
  scale_colour_brewer(palette = "Set3")
```

### A continuous colour scale

```{r}
ggplot(metabric, mapping = aes(x=GATA3, y=ESR1)) +
  geom_point(mapping = aes(colour=Nottingham_prognostic_index))
```

```{r}
ggplot(metabric, mapping = aes(x=GATA3, y=ESR1)) +
  geom_point(mapping = aes(colour=Nottingham_prognostic_index)) +
  scale_colour_gradient(low="pink", high="green")
```








