---
title: "Assignment 2"
---

---

### Working with vectors, data frames and tibbles

**1. How many cars in the `mtcars` data set have 8 cylinders?**

Write an R instruction to calculate this.

```{r}

```

**2. Print out a subset of the `starwars` tibble containing just the characters with blond hair and blue eyes**

```{r}

```

**3. Convert the values for the heights of characters in the `starwars` tibble from centimetres to metres**

```{r}

```

What happened to the type of the height column?

Convert the height column back to metres again and see if you can work out how
to change it back to its original type.

```{r}

```

---

### Simulation: rolling 6-sided dice

**4. Use the `sample()` function to simulate the rolling of a 6-sided die**

If you look at the help page for `sample` you'll find there are two functions
with the same name, one in the `base` package and the other in the `dplyr`
package. The `sample` function in the `base` package is the one you need for
this exercise.

Count the number of sixes you get for 1000 rolls?

```{r}

```

What number would you expect to get? Calculate this in R.

```{r}

```

Re-run the simulation a few times by clicking on the green 'Run current chunk'
button for the code you've written. What do you notice?

Re-run the simulation with a much larger sample size. Check that the number of
sixes you obtain is much closer to the expected value.

```{r}

```

**5. Simulate the rolling of a weighted six-sided die**

Imagine we're running a casino and we have some weighted dice in which the
probability of throwing a six is a bit lower than it should be if all is fair
and square. Let's say the probability of a throwing a six is 0.1 while the
other 5 numbers all have equal probability.

Calculate the probability for throwing each of the other numbers and then,
referring again at the help page for `sample()`, re-run the simulation using
these probabilities. See how many sixes you get.

```{r}

```

*Hint:* use the `rep()` function to save some typing.

---

### Clinical cancer data

**6. How many samples in the METABRIC dataset have high cellularity but have no recorded classification with the 3-gene classifier?**

```{r}

```

*Hint:* this is actually quite tricky -- you may want to check for missing
values in the cellularity column as well as the column for the 3-gene
classifier.

---

### Working with data from an Excel spreadsheet

**7. Read into R one of your own Excel spreadsheets and explore the data using `summary()`, `table()` and/or `plot()`**

We haven't yet showed how to read Excel spreadsheets into R. There are more than
package providing this capability but for this exercise use the `readxl` package
from the tidyverse.

`readxl` should have been installed as part of the tidyverse but if you find
that's not the case, you will have to install it separately.

`readxl` is not loaded as one of the core tidyverse packages when you run
`library(tidyverse)` so you'll have to load this separately.

If you don't have an Excel spreadsheet to hand, you can use the clinical
annotation data for the NCI-60 cell line data set (https://bioinformatics-core-shared-training.github.io/r-intro/data/cellline_nci60_clinical_data.xlsx).

```{r}

```

---

### Create a report for your assignment

Click on the 'Knit' menu at the top of this file and select either whichever
option you prefer to create an HTML, PDF or Word document version of your
assignment. This will run all the code chunks and "knit" the resulting
results with the surrounding text to produce a report.

#### Additional markdown features

**_Emphasis_**

Can use either underscores or asterisks and combination of both for both bold
and italic emphasis.

```
*italic*   **bold**
_italic_   __bold__
```

**_Inline Code_**

Use of backticks (`).

```
`mean(starwar$height)`
```

**_Horizontal Rule_**

Three of more astisks or dashes.

```
*******

-------
```

**_Plain Code Blocks_**

Plain code blocks are displayed in a fixed-width font but not evaluated.

```
This text is displayed verbatim / preformatted
```

Have a look at examples of these in the text parts of this R markdown document
and how these are rendered when you run the knit operation.

For more features, see the help page in RStudio that can be accessed from the
the help menu: **Help > Markdown Quick Reference**

---
