---
title: "Assignment 2"
---

**1. How many cars in the `mtcars` data set have 8 cylinders?**

Write an R instruction to calculate this.

```{r}

```

**2. Print out a subset of the `starwars` data frame containing just the characters with blond hair and blue eyes**

```{r}

```

**3. Use the `sample()` function to simulate the rolling of a 6-sided die**

If you look at the help page for `sample` you'll find there are two functions
with the same name, one in the `base` package and the other in the `dplyr`
package. The `sample` function in the `base` package is the one you need for
this exercise.

Count the number of sixes you get for 1000 rolls?

```{r}

```

What number would you expect to get? Calculate this in R.

```{r}

```

Re-run the simulation a few times by clicking on the green 'Run current chunk'
button for the code you've written. What do you notice?

Re-run the simulation with a much larger sample size. Check that the number of
sixes you obtain is much closer to the expected value.

```{r}

```

**4. Simulate the rolling of a weighted six-sided die?**

Imagine we're running a casino and we have some weighted dice in which the
probability of throwing a six is a bit lower than it should be if all is fair
and square. Let's say the probability of a throwing a six is 0.1 while the
other 5 numbers all have equal probability.

Calculate the probability for throwing each of the other numbers and then,
referring again at the help page for `sample()`, re-run the simulation using
these probabilities. See how many sixes you get.

```{r}

```

*Hint:* use the `rep()` function to save some typing.

**5. How many samples in the METABRIC dataset have high cellularity but have no recorded classification with the 3-gene classifier?**

```{r}

```

*Hint:* this is actually quite tricky -- you may want to check for missing
values in the cellularity column as well as the column for the 3-gene
classifier.

**6. Read into R one of your own Excel spreadsheets and explore the data using `summary()`, `table()` and/or `plot()`**

We haven't yet showed how to read Excel spreadsheets into R. There are more than
package providing this capability but for this exercise use the `readxl` package
from the tidyverse.

`readxl` should have been installed as part of the tidyverse but if you find
that's not the case, you will have to install it separately.

`readxl` is not loaded as one of the core tidyverse packages when you run
`library(tidyverse)` so you'll have to load this separately.

If you don't have an Excel spreadsheet to hand, you can use the clinical
annotation data for the NCI-60 cell line data set (https://bioinformatics-core-shared-training.github.io/r-intro/data/cellline_nci60_clinical_data.xlsx).

```{r}

```

